---
title: "accidents"
author: "Xuan Hung Ho"
date: "2023-08-25"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#install.packages(c("dplyr", "ggplot2")
library(dplyr)
library(tidyr)
library(ggplot2)
library(nnet)
```


```{r load data}
accidents = read.csv('../../data/clean/ACCIDENT.csv', header = TRUE, 
                     na.strings = c("", "NA", "Unknown"))

summary(accidents)
```

```{r}
accidents$ACCIDENT_TYPE <- as.factor(accidents$ACCIDENT_TYPE)

accidents$DAY_OF_WEEK <- as.factor(accidents$DAY_OF_WEEK)
accidents$Day.Week.Description <- as.factor(accidents$Day.Week.Description)

accidents$ACCIDENT_TYPE <- as.factor(accidents$ACCIDENT_TYPE)
accidents$Accident.Type.Desc <- as.factor(accidents$Accident.Type.Desc)

accidents$DCA_CODE <- as.factor(accidents$DCA_CODE)
accidents$DCA.Description <- as.factor(accidents$DCA.Description)

accidents$NODE_ID <- as.factor(accidents$NODE_ID)

accidents$LIGHT_CONDITION <- as.factor(accidents$LIGHT_CONDITION)
accidents$Light.Condition.Desc <- as.factor(accidents$Light.Condition.Desc)

accidents$ROAD_GEOMETRY <- as.factor(accidents$ROAD_GEOMETRY)
accidents$Road.Geometry.Desc <- as.factor(accidents$Road.Geometry.Desc)

accidents$SEVERITY <- factor(accidents$SEVERITY, ordered = TRUE, 
                                levels = c(1, 2, 3, 4))

accidents$SPEED_ZONE <- as.factor(accidents$SPEED_ZONE)
```

## 1. Univariate analysis

```{r}
ggplot(accidents, aes(Day.Week.Description)) + 
  geom_bar(width = 0.5) +
  theme_minimal() +
  labs(title="Distribution of Day of week", x = NULL, y = NULL)
```

It can be seen that accidents tend to happen more during Thursday and Friday, but the differences among days of week are not significant.

```{r}
ggplot(accidents, aes(SPEED_ZONE)) + 
  geom_bar() +
  theme_minimal() +
  labs(title="Distribution of Speed Zone", x = "Speed zone", y = NULL)
```

From this graph we can see there are couples of outliers. We will remove them before procceding further.

```{r}
accidents = accidents[as.numeric(as.character(accidents$SPEED_ZONE)) < 200, ]
accidents$SPEED_ZONE <- factor(accidents$SPEED_ZONE, ordered = TRUE, 
                                levels = c(30, 40, 50, 60, 70, 
                                           75, 80, 90, 100, 110))

ggplot(accidents, aes(SPEED_ZONE)) + 
  geom_bar() +
  theme_minimal() +
  labs(title="Distribution of Speed Zone", x = "Speed zone", y="Count")
```
We observe that most of the accidents happened in the 60km/h zone, which we are not expected.

```{r}
ggplot(accidents, aes(Road.Geometry.Desc)) + 
  geom_bar() +
  theme_minimal() +
  labs(title="Distribution of Geometry design of roads", 
       x = NULL, y = NULL) +
  theme(axis.text.x = element_text(angle = 30, vjust = 0.5))
```

We can conclude that more accidents took places on the road, not at intersections.

```{r}
ggplot(accidents, aes(Light.Condition.Desc)) + 
  geom_bar() +
  theme_minimal() +
  labs(title="Distribution of Light condition during accidents", 
       x = NULL, y = NULL) +
  theme(axis.text.x = element_text(angle = 30, vjust = 0.5))
```

Most of the accidents were recorded during day.

```{r}
ggplot(accidents, aes(Accident.Type.Desc)) + 
  geom_bar() +
  theme_minimal() +
  labs(title="Distribution of accident types", 
       x = NULL, y = NULL) +
  theme(axis.text.x = element_text(angle = 30, hjust = 0.7, vjust = 0.9))
```

Unsurprisingly, most of the accidents are between vehicles.

```{r}
ggplot(accidents, aes(SEVERITY)) + 
  geom_bar(width = 0.5) +
  theme_minimal() +
  labs(title="Distribution of Severity", x = NULL, y = NULL)
```

## 2. Bivariate analysis

We want to display the percentage of cases within each speed zone against the severity of crashes.

```{r}
aggregated_data <- accidents %>%
  group_by(SPEED_ZONE, SEVERITY) %>%
  tally() %>%
  group_by(SPEED_ZONE) %>%
  mutate(Percentage = n/sum(n) * 100)

ggplot(aggregated_data, aes(x = SPEED_ZONE, y = SEVERITY, fill = Percentage)) +
  geom_tile() + 
  scale_fill_gradient(low="white", high="red", name="% of Cases") + 
  labs(title="Heatmap of Speed Zones vs. Severity Percentage", 
       x="Speed Zone", y="Severity") + 
  theme_minimal()
```

From the heap map, we can say that in speed zone of 30 to 60km/h, there is higher chance that accidents is more severe.

## 3. Model fitting

Now we will try to fit multinom model, with "Light condition", "Speed zone" and "Road geometry" as predictors, and "Severity" as response variable. Since the response are multi-level factor, fitting multinom models are justifiable.

```{r}
add_model = multinom(SEVERITY ~ Light.Condition.Desc + SPEED_ZONE +
                  Road.Geometry.Desc, data = accidents)

interactive_model = multinom(SEVERITY ~ (Light.Condition.Desc + SPEED_ZONE +
                          Road.Geometry.Desc)^2, data = accidents)

anova(add_model, interactive_model, test = "Chi")
```
Since p-value equals 1, we have no evidence against the null hypothesis, which is the addictive model, i.e., we do not need interaction terms for the model.

```{r}
print(tidy(add_model, conf.int= TRUE, conf.level = 0.95, exponentiate = T), n = 100)
```
Since all p-values are relatively small (smaller than 0.05), we can say that all the variables in the model are significant.

```{r}
write.csv(accidents, '../../data/clean/ACCIDENT.csv', row.names = FALSE)
```

