---
title: "persons"
author: "Xuan Hung Ho"
date: "2023-08-27"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#install.packages(c("dplyr", "ggplot2")
library(dplyr)
library(tidyr)
library(tidyr)
library(ggplot2)
library(nnet)
```


```{r load data}
persons = read.csv('../../data/clean/PERSON.csv', 
                   header = TRUE)
```

## 1. Persons analysis

### 1.1. Univariate analysis
```{r}
summary(persons)
```


```{r}
persons$SEX <- as.factor(persons$SEX)

persons$Age.Group <- as.factor(persons$Age.Group)

persons$INJ_LEVEL <- factor(persons$INJ_LEVEL, ordered = TRUE, 
                                levels = c(9, 4, 3, 2, 1))

persons$Inj.Level.Desc <- factor(persons$Inj.Level.Desc, ordered = TRUE, 
                                levels = c('Unknown', 'Not injured', 
                                           'Other injury', 'Serious injury', 
                                           'Fatality'
                                           ))

persons$SEATING_POSITION <- as.factor(persons$SEATING_POSITION)

persons$ROAD_USER_TYPE <- as.factor(persons$ROAD_USER_TYPE)

persons$Road.User.Type.Desc <- as.factor(persons$Road.User.Type.Desc)
```

```{r}
ggplot(persons %>% filter(!is.na(AGE)), aes(AGE)) + 
  geom_bar(width = 0.5) +
  theme_minimal() +
  theme(axis.text.x = element_text()) +
  scale_x_continuous(breaks=seq(0,109,5)) +
  labs(title="Distribution by Age", x = NULL, y = NULL)
```

As it can be seen from the chart, people at the age of 18 are most likely to be involved in accidents. 

However, if we take a look at the problem from the age group point of view, the distribution looks differently:

```{r}
ggplot(persons, aes(Age.Group)) + 
  geom_bar(width = 0.5) +
  theme_minimal() +
  theme(axis.text.x = element_text()) +
  labs(title="Distribution by Age group", x = NULL, y = NULL)
```

We will take a deeper look this problem later, when we try to determine their roles in the accident, for example, drivers or passengers.

```{r}
summary(persons %>%
          filter(!is.na(HELMET_BELT_WORN)) %>% 
          select(HELMET_BELT_WORN)
  )
```

The variable 'HELMET_BELT_WORN' is supposed to be a boolean variable indicating whether the person invovled in the accidents was wearing seat belt/ helmet or not, but in the data set it is unjustifiable. Hence, we cannot use it to analyse/ build prediction model.

```{r}
ggplot(persons, aes(Road.User.Type.Desc)) + 
  geom_bar(width = 0.5) +
  theme_minimal() +
  theme(axis.text.x = element_text()) +
  labs(title="Distribution by User type", x = NULL, y = NULL)
```
```{r}
ggplot(persons, aes(SEATING_POSITION)) + 
  geom_bar(width = 0.5) +
  theme_minimal() +
  theme(axis.text.x = element_text()) +
  labs(title="Distribution by Seating position", x = NULL, y = NULL)
```

We cannot give a concrete conclusion from this graph, since there are chances that most of the drivers in the accidents might travel alone.
