---
title: "persons"
author: "Xuan Hung Ho"
date: "2023-08-27"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
library(MASS)
```


```{r load data}
persons = read.csv('../../data/clean/PERSON.csv', 
                   header = TRUE,
                   na.strings = c("", "NA", "Unknown"))

summary(persons)
```

```{r}
persons$SEX <- as.factor(persons$SEX)

persons$Age.Group <- factor(persons$Age.Group, ordered = TRUE,
                            levels = c("0-4", "5-12", "13-15", "16-17", "18-21", 
                                       "22-25", "26-29", "30-39", "40-49", 
                                       "50-59", "60-64", "65-69", "70+"))

persons$INJ_LEVEL <- factor(persons$INJ_LEVEL, ordered = TRUE,
                                levels = c(9, 4, 3, 2, 1))

persons$Inj.Level.Desc <- factor(persons$Inj.Level.Desc, ordered = TRUE, 
                                levels = c('Not injured', 
                                           'Other injury', 'Serious injury', 
                                           'Fatality'
                                           ))

persons$SEATING_POSITION <- as.factor(persons$SEATING_POSITION)
persons$SEATING_POSITION <- relevel(persons$SEATING_POSITION, ref = "D")

persons$ROAD_USER_TYPE <- as.factor(persons$ROAD_USER_TYPE)

persons$Road.User.Type.Desc <- as.factor(persons$Road.User.Type.Desc)

persons$HELMET_BELT_WORN <- as.factor(persons$HELMET_BELT_WORN)
```

## 1. Univariate analysis

```{r}
ggplot(persons, aes(SEX)) + 
  geom_bar(width = 0.5) +
  theme_minimal() +
  theme(axis.text.x = element_text()) +
  labs(title="Distribution by Gender", x = NULL, y = NULL)
```

More men are involved in accidents comparing to women.

```{r}
persons_filtered <- persons %>% filter(!is.na(AGE))
ggplot(persons_filtered, aes(AGE)) + 
  geom_bar(width = 0.5) +
  theme_minimal() +
  theme(axis.text.x = element_text()) +
  scale_x_continuous(breaks=seq(0,109,5)) +
  labs(title="Distribution by Age", x = NULL, y = NULL)
```
As it can be seen from the chart, people at the age of 18 are most likely to be involved in accidents. 

However, if we take a look at the problem from the age group point of view, the distribution looks differently:

```{r}
ggplot(persons_filtered, aes(Age.Group)) + 
  geom_bar(width = 0.5) +
  theme_minimal() +
  theme(axis.text.x = element_text()) +
  labs(title="Distribution by Age group", x = NULL, y = NULL)
```

We will take a deeper look this problem later, when we try to determine their roles in the accident, for example, drivers or passengers.

```{r}
ggplot(persons %>% filter(!is.na(HELMET_BELT_WORN) & HELMET_BELT_WORN != 9), 
       aes(HELMET_BELT_WORN)) + 
  geom_bar(width = 0.5) +
  theme_minimal() +
  theme(axis.text.x = element_text()) +
  labs(title="Distribution by Helmet/Seat belt worn", x = NULL, y = NULL)
```
```{r}
# Code Description 1 Seatbelt worn 2 Seatbelt not worn 3 Child restraint worn 4 Child restraint not worn 5 Seatbelt/restraint not fitted 6 Crash helmet worn 7 Crash helmet not worn 8 Not appropriate 9 Not known
# Source: https://data.vicroads.vic.gov.au/Metadata/Crash%20Stats%20-%20Data%20Extract%20-%20Open%20Data.html

persons <- persons %>%
  mutate(HELMET_BELT_WORN_PROPERLY = case_when(
    HELMET_BELT_WORN %in% c(1, 3, 6) ~ TRUE,
    HELMET_BELT_WORN == 9 ~ NA,
    .default = FALSE
  ))

persons$HELMET_BELT_WORN_PROPERLY <-
  as.factor(persons$HELMET_BELT_WORN_PROPERLY)

write.csv(persons, '../../data/clean/PERSON.csv', row.names = FALSE)
```


```{r}
ggplot(persons, aes(Road.User.Type.Desc)) + 
  geom_bar(width = 0.5) +
  theme_minimal() +
  theme(axis.text.x = element_text()) +
  labs(title="Distribution by User type", x = NULL, y = NULL)
```

```{r}
# Code Description CF Centre-front CR Centre-rear D Driver or rider LF Left-front LR Left-rear NA Not applicable NK Not known OR Other-rear PL Pillion passenger PS Motorcycle sidecar passenger RR Right-rear

ggplot(persons, aes(SEATING_POSITION)) + 
  geom_bar(width = 0.5) +
  theme_minimal() +
  theme(axis.text.x = element_text()) +
  labs(title="Distribution by Seating position", x = NULL, y = NULL)
```

We cannot give a concrete conclusion from this graph since there are chances that most of drivers in the accidents might travel alone. We will dive deeper into this problem later.

## 2. Bivariate analysis

We want to display the age of people against the injured severity.

```{r}
ggplot(persons_filtered, aes(x = AGE, y = Inj.Level.Desc, fill = SEX)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title="Boxplot of Age group by Injuried level", 
       x="Age", y="Injuried level")
```

We will try to test the correlation between person's seating position and injurious level:

```{r}
persons_filtered <- persons %>% 
    filter(!is.na(Inj.Level.Desc) & SEATING_POSITION != "PS" & 
             !is.na(SEATING_POSITION))

persons_filtered$Inj.Level.Desc <- factor(persons_filtered$Inj.Level.Desc)
persons_filtered$SEATING_POSITION <- factor(persons_filtered$SEATING_POSITION)

(chisq.test(
  table(persons_filtered$SEATING_POSITION, persons_filtered$Inj.Level.Desc),
  simulate.p.value = TRUE
))
```
We can see that there is huge gap between numbers of injuries persons among different seating positions, so the test result might not be correct. But from this result, it is evident that there is a correlation between seating positions and the level of injurious.

To consider all cases where only cars are involved, we first filter the data set, find all vehicles that have more than 2 people.

```{r}
persons_filtered_accident_no <- persons_filtered %>%
  filter(Road.User.Type.Desc %in% c("Drivers",
                                    "Passengers")) %>%
  group_by(ACCIDENT_NO, VEHICLE_ID) %>%
  count() %>%
  filter(n >= 2)

persons_filtered <- persons_filtered %>%
  filter(ACCIDENT_NO %in% persons_filtered_accident_no$ACCIDENT_NO)
```


```{r}
(chisq.test(
  table(persons_filtered$SEATING_POSITION, persons_filtered$Inj.Level.Desc),
  simulate.p.value = TRUE
))
```
With p-value < 0.05, we can say that with 95% confidence level, the person's seating in cars is strongly related to the level of injurious.

## 3. Model fitting

We will try to fit a ordinal regression model to see if there is a connection between seating positions, age group against injurious level.

```{r}
persons_filtered <- persons_filtered %>%
  filter(!is.na(Age.Group) & !is.na(HELMET_BELT_WORN_PROPERLY)) %>%
  filter(!is.na(SEX))

add_model = polr(Inj.Level.Desc ~ SEATING_POSITION + Age.Group + 
                   HELMET_BELT_WORN_PROPERLY + SEX,
                 data = persons_filtered, Hess = TRUE)

summary(add_model)
```

```{r}
ctable <- coef(summary(add_model))
p <- pnorm(abs(ctable[, "t value"]), lower.tail = FALSE) * 2

(ctable <- cbind(ctable, "p value" = p))
```
From the p-values, we can say that all the variables in the model are significant.

```{r}
exp(coef(add_model))
```

We can interpret the table above as follows:

- For individuals within the same age group and seat belt/helmet status, if they are seated in the 'Passenger Left' position, the odds of moving from lower level of injurious to higher level (for example, having have non injury accident compared to other injuries, or from being seriously injured to fatal accident) in other seating positions in the car are relatively the same. \
- For people seated in the same position in a car and have same seat belt/helmet status, the odds of sustaining injuries for those in the age groups 5-12 and 16-17 are 2.46 and 1.77 times higher, respectively, compared to people in the age group 0-4. For other age groups, the odds are nearly the same. \
- For people who worn seat belt/ helmet properly, the odds of being injured are 46% (1-0.537) lower than those who did not wear seat belt/ helmet properly.
- The odds ratio for 'Sex Male' is 0.6084, which means that the odds of a higher severity accident are approximately 1.64 (1/0.684) times lower if the driver is a man compared to female drivers.

