---
title: "vehicles"
author: "Xuan Hung Ho"
date: "2023-08-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(tidyr)
library(ggplot2)
library(nnet)
```


```{r load data}
vehicles = read.csv('../../data/clean/VEHICLE.csv', header = TRUE, 
                     na.strings = c("", "NA", "Unknown"))

summary(vehicles)
```

```{r}
vehicles$ROAD_SURFACE_TYPE <- as.factor(vehicles$ROAD_SURFACE_TYPE)

vehicles$Road.Surface.Type.Desc <- as.factor(vehicles$Road.Surface.Type.Desc)

vehicles$VEHICLE_BODY_STYLE <- as.factor(vehicles$VEHICLE_BODY_STYLE)

vehicles$VEHICLE_MAKE <- as.factor(vehicles$VEHICLE_MAKE)
vehicles$VEHICLE_MODEL <- as.factor(vehicles$VEHICLE_MODEL)

vehicles$VEHICLE_TYPE <- as.factor(vehicles$VEHICLE_TYPE)
vehicles$Vehicle.Type.Desc <- as.factor(vehicles$Vehicle.Type.Desc)

vehicles$FUEL_TYPE <- as.factor(vehicles$FUEL_TYPE)

vehicles$LEVEL_OF_DAMAGE <- factor(vehicles$LEVEL_OF_DAMAGE, ordered = TRUE, 
                                levels = c(1, 2, 3, 4, 5, 6, 7, 9))

vehicles$TRAFFIC_CONTROL <- as.factor(vehicles$TRAFFIC_CONTROL)
vehicles$Traffic.Control.Desc <- as.factor(vehicles$Traffic.Control.Desc)

vehicles$CAUGHT_FIRE <- as.factor(vehicles$CAUGHT_FIRE)
```

## 1. Univariate analysis

```{r}
vehicles_filtered <- vehicles %>%
  filter(!is.na(VEHICLE_YEAR_MANUF))

ggplot(vehicles_filtered, aes(VEHICLE_YEAR_MANUF)) + 
  geom_density(alpha=0.25) +
  xlim(1960, 2023)
```

```{r}
ggplot(vehicles, aes(Road.Surface.Type.Desc)) + 
  geom_bar() +
  theme_minimal() +
  labs(title="Distribution of Road surface type", x = NULL, y = NULL)
```

```{r}
ggplot(vehicles %>% filter(!is.na(Vehicle.Type.Desc)), aes(Vehicle.Type.Desc)) + 
  geom_bar() +
  theme_minimal() +
  labs(title="Distribution of Vehicle type", x = NULL, y = NULL) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```
Most of vehicles involved in accidents are cars.

```{r}
ggplot(vehicles %>% filter(!is.na(FUEL_TYPE)), aes(FUEL_TYPE)) + 
  geom_bar() +
  theme_minimal() +
  labs(title="Distribution of Fuel type", x = NULL, y = NULL)
```

Most of the vehicles involved in accidents were running on Petrol fuel

## 2. Bivariate analysis

We want to display the fuel type against whether the vehicles caught fire.

```{r}
vehicles_filtered <- vehicles %>% 
  filter(!is.na(FUEL_TYPE) & !is.na(CAUGHT_FIRE))

aggregated_data <- vehicles_filtered %>%
  group_by(FUEL_TYPE, CAUGHT_FIRE) %>%
  tally()

ggplot(aggregated_data, aes(x = FUEL_TYPE, y = CAUGHT_FIRE, fill = n)) +
  geom_tile() + 
  scale_fill_gradient(low="white", high="red", name="Count") + 
  labs(title="Heat map of Fuel type vs. Caught Fire", 
       x="Fuel type", y="Caught Fire") + 
  theme_minimal()
```

```{r}
table(vehicles$CAUGHT_FIRE)
```
Since we only record 1187 cases of catching fire, we might not be able to use it to analysis whether there are correlation between fuel type and whether a vehicle catching fire or not.
