---
title: "modelling"
author: "Xuan Hung Ho"
date: "2023-08-29"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
library(MASS)
```


```{r load data}
na.strings <- c("", "NA", "Unknown")

accidents = read.csv('../data/clean/ACCIDENT.csv', header = TRUE, 
                     na.strings = na.strings)

atmospheric_conds = read.csv('../data/clean/ATMOSPHERIC_COND.csv', 
                             header = TRUE, 
                             na.strings = na.strings)

persons = read.csv('../data/clean/PERSON.csv', 
                   header = TRUE,
                   na.strings = na.strings)

road_surface_conds = read.csv('../data/clean/ROAD_SURFACE_COND.csv', 
                              header = TRUE, 
                              na.strings = na.strings)

vehicles = read.csv('../data/clean/VEHICLE.csv', header = TRUE, 
                     na.strings = na.strings)
```

```{r}
df <- accidents %>% 
  inner_join(persons, by="ACCIDENT_NO") %>%
  inner_join(atmospheric_conds, by = "ACCIDENT_NO") %>%
  inner_join(road_surface_conds, by="ACCIDENT_NO") %>%
  inner_join(vehicles, by = c('ACCIDENT_NO', 'VEHICLE_ID'))


```

```{r}
# Accidents
df$ACCIDENT_TYPE <- as.factor(df$ACCIDENT_TYPE)

df$Accident.Type.Desc <- as.factor(df$Accident.Type.Desc)

df$Light.Condition.Desc <- as.factor(df$Light.Condition.Desc)
df$Light.Condition.Desc <- relevel(df$Light.Condition.Desc,
                                          ref = 'Day')

df$Road.Geometry.Desc <- as.factor(df$Road.Geometry.Desc)
df$Road.Geometry.Desc <- relevel(df$Road.Geometry.Desc, 
                                        ref = 'Cross intersection')

df$SEVERITY <- factor(df$SEVERITY, ordered = TRUE, 
                             levels = c("Non injury accident", 
                                        "Other injury accident", 
                                        "Serious injury accident", 
                                        "Fatal accident"))

df$SPEED_ZONE <- as.factor(df$SPEED_ZONE)
df$SPEED_ZONE <- relevel(df$SPEED_ZONE, ref = "60")

# Atmospheric conditions
df$Atmosph.Cond.Desc <- as.factor(df$Atmosph.Cond.Desc)

# Persons
df$SEX <- as.factor(df$SEX)

df$Age.Group <- factor(df$Age.Group, ordered = TRUE,
                            levels = c("0-4", "5-12", "13-15", "16-17", "18-21", 
                                       "22-25", "26-29", "30-39", "40-49", 
                                       "50-59", "60-64", "65-69", "70+"))

df$Inj.Level.Desc <- factor(df$Inj.Level.Desc, ordered = TRUE, 
                                levels = c('Not injured', 
                                           'Other injury', 'Serious injury', 
                                           'Fatality'
                                           ))

df$SEATING_POSITION <- as.factor(df$SEATING_POSITION)
df$SEATING_POSITION <- relevel(df$SEATING_POSITION, ref = "D")

df$ROAD_USER_TYPE <- as.factor(df$ROAD_USER_TYPE)

df$Road.User.Type.Desc <- as.factor(df$Road.User.Type.Desc)

df$HELMET_BELT_WORN_PROPERLY <-
  as.factor(df$HELMET_BELT_WORN_PROPERLY)

# Road surface condition
df$Surface.Cond.Desc <- 
  as.factor(df$Surface.Cond.Desc)

# Vehicles
df$Road.Surface.Type.Desc <- as.factor(df$Road.Surface.Type.Desc)

df$VEHICLE_BODY_STYLE <- as.factor(df$VEHICLE_BODY_STYLE)

df$Vehicle.Type.Desc <- as.factor(df$Vehicle.Type.Desc)
```
## 1. Factor analysis

In this section we will try to answer the question: Which factors are likely to be connected to severe accidents?

```{r}
df <- df %>%
  filter(!is.na(SEVERITY)) %>%
  filter(!is.na(Light.Condition.Desc)) %>%
  filter(!is.na(Atmosph.Cond.Desc)) %>%
  filter(!is.na(Surface.Cond.Desc)) %>%
  filter(!is.na(Road.Surface.Type.Desc)) %>%
  filter(!is.na(SPEED_ZONE)) %>%
  filter(!is.na(Road.Geometry.Desc)) %>%
  filter(!is.na(SEX)) %>%
  filter(!is.na(VEHICLE_WEIGHT))

model_data <- df %>%
  dplyr::select(SEVERITY, Light.Condition.Desc, Road.Surface.Type.Desc, 
                SPEED_ZONE, Road.Geometry.Desc, VEHICLE_WEIGHT, 
                Surface.Cond.Desc, Atmosph.Cond.Desc, SEX) %>%
  distinct(SEVERITY, Light.Condition.Desc, Road.Surface.Type.Desc, 
           SPEED_ZONE, Road.Geometry.Desc, VEHICLE_WEIGHT, 
           Surface.Cond.Desc, Atmosph.Cond.Desc, SEX,
           .keep_all = TRUE)

model_data$VEHICLE_WEIGHT <- model_data$VEHICLE_WEIGHT / 1000

model <- polr(SEVERITY ~ Light.Condition.Desc + VEHICLE_WEIGHT + SPEED_ZONE + 
                SEX + Surface.Cond.Desc + Atmosph.Cond.Desc + 
                Road.Surface.Type.Desc + Road.Geometry.Desc,
              data = model_data, Hess = TRUE)
```
It seems like one of the variables has caused the design matrix to be not full rank. We will try to remove "Road.Geometry.Desc":
```{r}
model <- polr(SEVERITY ~ Light.Condition.Desc + VEHICLE_WEIGHT + SPEED_ZONE + 
                SEX + Surface.Cond.Desc + Atmosph.Cond.Desc + 
                Road.Surface.Type.Desc,
              data = model_data, Hess = TRUE)
summary(model)
```
The design matrix is now full rank. Now we perform _step_ function to obtain the model with the lowest AIC score:

```{r}
model <- step(model)
```
Seems like _Atmosph.Cond.Desc_ does not contribute significantly to the model. We can verify all the other variables are significant by computing p-values, i.e., Wald test.

```{r}
ctable <- coef(summary(model))
p <- pnorm(abs(ctable[, "t value"]), lower.tail = FALSE) * 2

(ctable <- cbind(ctable, "p value" = p))
```
```{r}
exp(coef(model))
```
The odds ratios represent the multiplicative change in the odds of being in a higher severity category for a one-unit increase in the predictor variable, holding the other variables constant.

For example, the odds ratio for 'Dark No street lights ' is 1.4744, which means that the odds of a higher severity accident are 1.47 times higher when it is dark with no street lights compared to accidents happened during day.

Similarly, the odds ratio for 'Speed Zone 100' is 1.5883, which means that the odds of a higher severity accident are approximately 1.59 times higher in a 100 speed zone compared to the Speed Zone 60.

The odds ratio for 'Sex Male' is 1.1389634, which means that the odds of a higher severity accident are approximately 1.14 times higher if the driver is a man compared to female drivers.

## 2. Time series analysis

